Fail Faster

### 2025-03-24 paper/paper.tex
CP

In both languages, our libraries preserve the semantics of the original: given the same random seed, generators written using \camlname or \scalaname produce exactly the same sequence of values as equivalent generators using \bq or ScalaCheck, respectively.
This semantic equivalence enables pointwise---as opposed to distributional---bug-finding comparisons, ensuring that if a \name{} generator finds bugs faster, this is attributable \textit{solely} to staging and \rand optimizations, not luck with seeds.

We evaluate the optimizations with a series of case studies in each
generator library, and show that the \name{} generators run faster than their unstaged equivalents.
Further, we show generators that run faster also find bugs faster
by running these case studies in the Etna
platform~\cite{etna}, which simulates real-world PBT usage \iflater\hg{Don't love this phrase, I'll
think more about how we should talk about Etna}\fi by injecting bugs into a system and
measuring how quickly various generators detect them.
\jwc{Beef up this paragraph, explain the results of our eval.}

In summary, we show that PBT generator DSLs incur significant performance costs across languages, and present a general technique for improving their efficiency while preserving their idiomatic abstractions. \jwc{needs a bit about RNG --- BCP thoughts?}
We make the following contributions:
\begin{enumerate}
    \item We identify two key sources of inefficiency in PBT generator libraries, which can significantly impact performance:
    abstraction overhead and \rand choice.
    \item We present \name{}, a technique for eliminates the abstraction overhead of generators. We implement \name{} in both OCaml and Scala 3, showcasing its cross-language applicability.
    \item We demonstrate that writing generators using \name{} and fast \rands{} yields significant performance improvements.
    We precisely quantify the impact of each optimization separately, and explain how they play distinct yet complementary roles in improving generator performance.
    Finally, we show that these performance improvements extend to improved bug-finding speed, finding some bugs as much as $13\times$ faster.
\end{enumerate}

In Section~\ref{section:motiv}, we introduce monadic generator DSLs, discuss how they work, and take an example-guided tour through some of the ways that
they are slower than is possible.
Then, in Section~\ref{section:faster-generators}, we explain multi-stage programming,
develop \name{} (using \camlname{} to illustrate), and discuss how we isolate the performance of \rand{} to benchmark their impact on bug-finding speed.
Section~\ref{section:eval} evaluates all of the previous work, demonstrating the generator performance improvements
and how they lead to much faster bug finding.

@@

With a sense of the inefficiences pervasive in generator libraries, we set out in this section to investigate solutions.
We begin with a quick tour through multi-stage programming, then incrementally build \camlname, layering on features and functionality.
Last, we present the controlled experiment we use to demonstrate how a faster \rand{} leads to failing faster.

@@

\paragraph{Uses of Staging in eDSL Construction}
One of the primary uses of staging is in embedded DSL construction \cite{sheard99, lms, trattdsl}.
\jwc{Cite racket macros for DSLs here:}
While embedded DSLs are powerful tools, they have a well-known drawback:
the functional abstractions used to build eDSLs tend to prevent compilers from generating efficient machine code \cn.
This is \emph{abstraction overhead} \cite{carrette05, moller20}: the layers of abstraction that
make eDSLs so pleasant to use are precisely what prevents them from being fast.
Indeed, the root causes of many of the performance issues we discovered in Section~\ref{section:motiv} are
not unique to generator DSLs, but pervasive across eDSLs.
In light of this issue, staging is often used as a lightweight compiler for DSLs. The compile-time evaluation stage
transforms the DSL code, eliminating abstractions to produce code that the host language compiler can generate fast machine code for.
This recipe has been used to great effect to stage eDSLs for stream processing \cite{moller20,strymonas}, parser combinators \cite{sspc, staged-parsers, krishnaswami19, flap},
and query processing \cite{rhyme}.

@@

\section{Related Work}

@@

\section{Conclusion \& Future Work}
In this paper, we identified, studied, and proposed potential solutions to two
important sources of inefficiency in PBT generator libraries: abstraction
overhead and choice of \rand{}.  In the future, we hope to continue to investigate and push the boundaries of
PBT generator performance.

>>>>

Both interventions preserve semantics on the nose: given the same
random seed, generators written using \camlname or \scalaname produce
exactly the same sequence of values as equivalent generators using \bq
or ScalaCheck.  This semantic equivalence enables pointwise---as
opposed to distributional---comparisons of bug-finding effectiveness,
ensuring that, if a \name{} generator finds bugs faster, this is
attributable \textit{solely} to the staging and \rand interventions,
not lucky choice of seeds.

We evaluate the optimizations with a series of case studies in each
generator library, and show that the \name{} generators run faster
than their unstaged equivalents.  Further, we show generators that run
faster also find bugs faster by running these case studies in the Etna
platform~\cite{etna}, which injects bugs into a system and measures
how quickly various generators detect them.  \jwc{Beef up this
  paragraph, explain the results of our eval.}  \todo{We precisely
  quantify the impact of each intervention and explain how they play
  distinct yet complementary roles in improving generator performance.
}

In summary, we show that PBT generator DSLs incur significant
performance costs across languages and present two interventions that
significantly improve their efficiency while preserving their
idiomatic style.  Concretely, we offer the following contributions:
\begin{enumerate}
\item We identify two key sources of inefficiency in PBT generator
libraries---abstraction overhead and choice of \rand---both of which
significantly impact performance (Section~\ref{section:motiv}).
\item We present \name{}, a staging technique that eliminates the
abstraction overhead of generators. We apply \name{} to standard
generator libraries in both OCaml and Scala 3, showcasing its
generality (Section~\ref{section:faster-generators}).
\item We demonstrate that writing generators using \name{} and fast
\rands{} yields substantial performance improvements, both separately
and in combination, and we show that these performance improvements
extend to significantly improved bug-finding speed (Section~\ref{section:eval}).
\end{enumerate}

Section~\ref{section:other-langs} discusses how the \name{}
technique could be applied to PBT libraries in other
languages---Racket, F\#, Haskell, and Rust.
%
Sections~\ref{section:related} and~\ref{section:concl} present related
and future work.

@@

With a sense of two main inefficiences in generator libraries, we set out in this section to investigate solutions.
We begin with a quick tour through multi-stage programming (\ref{xx}),
then incrementally build \camlname, layering on features and
functionality  (\ref{xx}).
Last, we present the controlled experiment we use to demonstrate how a
faster \rand{} leads to failing faster  (\ref{xx}).

@@

\paragraph{Staging eDSLs}
One of the primary uses of staging is in embedded DSL construction \cite{sheard99, lms, trattdsl}.
\jwc{Cite racket macros for DSLs here:}
While embedded DSLs are powerful tools, they have a well-known drawback:
the functional abstractions used to build eDSLs tend to prevent compilers from generating efficient machine code \cn.
This is \emph{abstraction overhead} \cite{carrette05, moller20}: the layers of abstraction that
make eDSLs so pleasant to use are precisely what prevents them from being fast.
Indeed, the root causes of many of the performance issues we discovered in Section~\ref{section:motiv} are
not unique to generator DSLs, but pervasive across eDSLs.
In light of this issue, staging is often used as a lightweight compiler for DSLs. The compile-time evaluation stage
transforms the DSL code, eliminating abstractions to produce code that the host language compiler can generate fast machine code for.
This recipe has been used to great effect to stage eDSLs for stream processing \cite{moller20,strymonas}, parser combinators \cite{sspc, staged-parsers, krishnaswami19, flap},
and query processing \cite{rhyme}.

@@

\section{Related Work}
\label{section:related}

@@

\section{Conclusion \& Future Work}
\label{section:concl}
In this paper, we identified, studied, and proposed potential solutions to two
important sources of inefficiency in PBT generator libraries: abstraction
overhead and choice of \rand{}.  In the future, we hope to continue to investigate and push the boundaries of
PBT generator performance.
